---
title: "Immunarch Analysis"
output: html_notebook
---

load libraries
```{r}
library("immunarch")
library("purrr")
library("stringr")
library(forcats)
library(data.table)
library(ggpubr)
library(scales)
library(plotly)
library(dplyr)
library(vegan)
library(ggplot2)
library(RColorBrewer)
library(ComplexHeatmap)
library(stats)
```

load repertoires
```{r}
# Load TCR repertoire data from PBMC samples in specified directory
lung <- repLoad("~/Desktop/TCR-PBMC")
```


low clones
```{r}
# Explore clone counts in the loaded data
clones <- repExplore(lung$data, .method = "clones")
# Arrange samples by clone count
blood_order = arrange(clones, Clones)
# Remove row names for cleaner output
rownames(blood_order) = c()
# Display top 8 entries
head(blood_order, 8)
```

omit small samples
```{r}
# Remove specific sample "MDA-4378-C" from the dataset
lung$data[which(names(lung$meta) %in% c("MDA-4378-C"))] <- NULL
```

barplot of sample frequency of each stage
```{r}
# Create dataframe counting frequency of disease stages for PBMC samples only
stage_frequency <- data.frame(value = toupper(lung$meta[lung$meta$`Tissue Source`=="PBMC",]$`Disease Stage`)) %>%
  count(value)
# Generate bar plot showing frequency by disease stage
ggplot(stage_frequency, aes(x = value, y = n, fill = value)) +
  geom_bar(stat = "identity") +                    # Create bars with actual counts
  labs(title="Sample frequency by disease stage", x = "Disease stage", y = "Frequency") + 
  scale_color_discrete(name="Stage", labels=c("I","II","III")) +  # Customize legend
  scale_fill_grey()                                # Use grey color scale for fill
```


subsampling for more accurate analysis
```{r}
# Downsample the repertoire data to normalize sequencing depth
sub <- repSample(.data = lung$data, .method = "downsample")
```

```{r}
# Save subsampled data and metadata for use in other scripts
saveRDS(sub, "sub-pbmc.Rdata")          # Save subsampled repertoire data
saveRDS(lung$meta, "meta.Rdata")        # Save metadata
```

```{r}
# Initialize empty lung object and reload saved data
lung <- {}
sub <- readRDS(file="sub-pbmc.Rdata")    # Load subsampled data
lung$meta <- readRDS(file="meta.Rdata")  # Load metadata
```

shannon diversity + richness index
```{r}
# Calculate Shannon diversity index for each subsampled repertoire
shannon_div <- lapply(sub, function(x) {diversity(x$Clones,index = "shannon")})
# Combine diversity indices with disease stage information
shannon_div <- as.matrix(shannon_div) %>% cbind(sapply(names(sub), function(x){lung$meta[lung$meta$Sample == x,]$`Disease Stage`}))

# Create dataframe for diversity metrics
df_div <- as.data.frame(shannon_div)
colnames(df_div) <- c("shan","stage")    # Rename columns
# Convert data types for plotting
df_div$stage <- as.factor(as.numeric(factor(unlist(df_div$stage))))  # Convert stage to factor
df_div$shan <- as.double(df_div$shan)    # Convert Shannon index to double
# Calculate richness (unique clonotypes) for subsampled and original data
df_div$rich <- as.double(lapply(sub, function(x) {specnumber(x$Clones)}))
df_div$presubrich <- as.double(lapply(lung$data, function(x) {specnumber(x$Clones)}))
```

richness barplot pre subsample
```{r}
# Plot richness index for original (pre-subsampled) data
ggplot(df_div, aes(x = rownames(df_div), y = presubrich)) +
  geom_bar(stat='identity') + 
  labs(title = "Richness index", x="Samples", y="Richness level") +
  theme_bw() + theme(axis.text.x=element_blank(),    # Remove x-axis labels for clarity
        axis.ticks.x=element_blank())
```

richness barplot
```{r}
# Plot richness index for subsampled data
ggplot(df_div, aes(x = rownames(df_div), y = rich)) +
  geom_bar(stat='identity') + 
  labs(title = "Richness index", x="Samples", y="Richness level") +
  theme_bw() + theme(axis.text.x=element_blank(),    # Remove x-axis labels for clarity
        axis.ticks.x=element_blank())
```


richness index by disease stage
```{r}
# Boxplot of richness by disease stage
ggplot(df_div, aes(x = stage, y = rich, colour = stage)) +
  geom_boxplot() + 
  scale_color_discrete(name="Stage", labels=c("I","II","III","H")) +  # Color by stage
  labs(title = "Richness index by disease stage", x="Disease stage", y="Richness level") +
  theme_bw() +
  stat_compare_means()    # Add statistical comparison (Kruskall-Wallis)
# Perform ANOVA with Tukey's post-hoc test
TukeyHSD(aov(rich~stage,data=df_div))
```

shannon diversity by disease stage
```{r}
# Boxplot of Shannon diversity by disease stage
ggplot(df_div, aes(x = stage, y = shan, colour = stage)) +
  geom_boxplot() + 
  scale_color_discrete(name="Stage", labels=c("I","II","III","H")) +  # Color by stage
  labs(title = "Shannon diversity index by disease stage", x="Disease stage", y="Diversity level") +
  theme_bw() +
  stat_compare_means() +  # Add statistical comparison (Kruskall-Wallis)
  stat_compare_means()
# Perform ANOVA with Tukey's post-hoc test
TukeyHSD(aov(shan~stage,data=df_div))
```


top clonal proportions by #health status# (obsolete) + disease stage, once with #healthy samples#, once without
```{r}
# Calculate top clonal proportions
clontop <- repClonality(sub, "top")
# Visualize clonal proportions by disease stage
vis(clontop, .by = "Disease Stage", .meta = lung$meta)
# Add clonal proportion data to dataframe
df_div$clonal10 <- clontop[,"10"]
df_div$clonal3000 <- clontop[,"3000"]
# Perform ANOVA tests for clonal proportions
TukeyHSD(aov(clonal10~stage,data=df_div))
TukeyHSD(aov(clonal3000~stage,data=df_div))
```

repetoire overlap analysis
```{r}
# Calculate repertoire overlap
public <- repOverlap(sub, .method = "public", .verbose = F, .col = "aa")    # Public clonotypes
morisita <- repOverlap(sub, .method = "morisita", .verbose = F, .col = "aa") # Morisita index
jaccard <- repOverlap(sub, .method = "jaccard", .verbose = F, .col = "aa")   # Jaccard index
```

```{r}
# Save overlap analysis results for future use
saveRDS(morisita, "morisita-pbmc.Rdata")
saveRDS(public, "public-pbmc.Rdata")
saveRDS(jaccard, "jaccard-pbmc.Rdata")
```
```{r}
# Load previously saved overlap analysis results
morisita <- readRDS(file="morisita-pbmc.Rdata")
public <- readRDS(file="public-pbmc.Rdata")
jaccard <- readRDS(file="jaccard-pbmc.Rdata")
```

display overlap analysis ordered by disease stage
```{r}
# Match sample indices between diversity dataframe and overlap matrices
indices <- match(rownames(df_div), rownames(morisita))
# Create metadata matrix with matching indices
metadata_matrix <- df_div[indices, ]
# Order metadata by disease stage
ordered_metadata <- metadata_matrix[order(metadata_matrix$stage), ]
# Reorder overlap matrices by disease stage
reordered_repOverlap_mor <- morisita[order(metadata_matrix$stage), order(metadata_matrix$stage)]
reordered_repOverlap_jac <- jaccard[order(metadata_matrix$stage), order(metadata_matrix$stage)]
reordered_repOverlap_pub <- public[order(metadata_matrix$stage), order(metadata_matrix$stage)]
# Create annotation list for disease stages
stage_anno_list <- setNames(as.character(ordered_metadata$stage), rownames(ordered_metadata))

# Generate heatmaps for each overlap metric
heat_mor <- Heatmap(reordered_repOverlap_mor, column_title = "Morisita overlap ordered by disease stage", cluster_rows = FALSE, cluster_columns = FALSE,
        top_annotation = HeatmapAnnotation(stage_top = stage_anno_list, col = list(stage_top = c("1" = "orange", "2" = "red", "3" = "brown")),
                                        annotation_legend_param = list(stage_top = list(labels = c("I", "II", "III")))),
        left_annotation = rowAnnotation(stage_left = stage_anno_list, col = list(stage_left = c("1" = "orange", "2" = "red", "3" = "brown")),
                                        annotation_legend_param = list(stage_left = list(labels = c("I", "II", "III")))))
heat_jac <- Heatmap(reordered_repOverlap_jac, column_title = "Jaccard overlap ordered by disease stage", cluster_rows = FALSE, cluster_columns = FALSE,
        top_annotation = HeatmapAnnotation(stage_top = stage_anno_list, col = list(stage_top = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_top = list(labels = c("I", "II", "III")))),
        left_annotation = rowAnnotation(stage_left = stage_anno_list, col = list(stage_left = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_left = list(labels = c("I", "II", "III")))))
heat_pub <- Heatmap(reordered_repOverlap_pub, column_title = "Public overlap ordered by disease stage", cluster_rows = FALSE, cluster_columns = FALSE,
        top_annotation = HeatmapAnnotation(stage_top = stage_anno_list, col = list(stage_top = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_top = list(labels = c("I", "II", "III")))),
        left_annotation = rowAnnotation(stage_left = stage_anno_list, col = list(stage_left = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_left = list(labels = c("I", "II", "III")))))
# Draw the heatmaps
draw(heat_mor)
draw(heat_jac)
draw(heat_pub)
```

display overlap analysis by clustering
```{r}
# Generate clustered heatmaps for overlap metrics
pub_clust <- Heatmap(public, column_title = "Public overlap ordered by clustering", 
        top_annotation = HeatmapAnnotation(stage_top = stage_anno_list, col = list(stage_top = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_top = list(labels = c("I", "II", "III")))),
        left_annotation = rowAnnotation(stage_left = stage_anno_list, col = list(stage_left = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_left = list(labels = c("I", "II", "III")))))
mor_clust <- Heatmap(morisita, column_title = "Morisita overlap ordered by clustering", 
        top_annotation = HeatmapAnnotation(stage_top = stage_anno_list, col = list(stage_top = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_top = list(labels = c("I", "II", "III")))),
        left_annotation = rowAnnotation(stage_left = stage_anno_list, col = list(stage_left = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_left = list(labels = c("I", "II", "III")))))
jac_clust <- Heatmap(jaccard, column_title = "Jaccard overlap ordered by clustering", 
        top_annotation = HeatmapAnnotation(stage_top = stage_anno_list, col = list(stage_top = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_top = list(labels = c("I", "II", "III")))),
        left_annotation = rowAnnotation(stage_left = stage_anno_list, col = list(stage_left = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_left = list(labels = c("I", "II", "III")))))
# Draw the clustered heatmaps
draw(pub_clust)
draw(mor_clust)
draw(jac_clust)
```

```{r}
mor_mds <- repOverlapAnalysis(morisita, "mds+kmeans", .k = 5) # 5
jac_mds <- repOverlapAnalysis(jaccard, "mds+kmeans", .k=7) # 7
pub_mds <- repOverlapAnalysis(public, "mds+kmeans", .k=6) # 6
vis(mor_mds, .by = "Disease Stage", .meta = lung$meta)
vis(jac_mds, .by = "Disease Stage", .meta = lung$meta)
vis(pub_mds, .by = "Disease Stage", .meta = lung$meta)
```

generate dataframes: features x observations. 
proportions of public clonotype sharing
```{r}
#lc_seqs <- pubRep(sub, .quant = "prop", .verbose = F, .col = "aa")
#tumor_seqs <- pubRep(sub_t, .quant = "prop", .verbose = F, .col = "aa")
df_seq_lc <- as.data.frame(pubRep(sub, .quant = "prop", .verbose = F, .col = "aa"))
```

remove na and sample column
```{r}
df_seq_lc[is.na(df_seq_lc)] = 0
#lc_in_t[is.na(lc_in_t)] = 0
df_seq_lc$Samples <- NULL
#df_seq_t_wos <- subset(lc_in_t, select = -c(Samples) )
```

transpose dataframe: observations x features
```{r}
tr_lc <- setNames(data.frame(t(df_seq_lc[,-1])), df_seq_lc[,1])
#tr_t <- setNames(data.frame(t(lc_in_t[,-1])), lc_in_t[,1])
```

check how many coulmns are in the df
```{r}
ncol(tr_lc)
```

perform pca on #healthy# (obsolete) + sick
```{r}
# Perform principal component analysis (PCA) on transposed data
pca_all_result <- prcomp(tr_lc, center = T)
```

plot pca
```{r}
# Plot PCA results colored by disease stage
ggplot(pca_all_result$x, aes(PC1,PC2,color=df_div$stage)) + 
  geom_point() + 
  scale_color_discrete(name="Stage", labels=c("I","II","III","H"))
```

```{r}
# Load F1 score results from CSV file
f1_res = read.csv("./f1_results.csv")
```

```{r}
# Bar plot comparing F1 scores across different ML models and matrices
ggplot(f1_res, aes(x=Model,y=f1.score,fill=Matrix)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  ggtitle("ML Classification algorithms evaluation with F1 score") + 
  labs(x= "ML model", y= "F1 Score") + 
  theme_bw()
```

```{r}
# Bar plot of F1 scores for threshold matrix, colored by model
ggplot(f1_res[f1_res$Matrix == "Threshold",], aes(x=Matrix,y=f1.score, fill=Model)) + 
  geom_bar(stat = "identity", position = "dodge") + 
  scale_fill_brewer(palette = "YlGn") +    # Use yellow-green color palette
  theme_bw() + 
  labs(y="F1 score", x="Threshold matrix") + 
  theme(axis.text.x=element_blank(),       # Remove x-axis labels
        axis.ticks.x=element_blank())
```
