---
title: "Immunarch Analysis"
output: html_notebook
---

load libraries
```{r}
library("immunarch")
library("purrr")
library("stringr")
library(forcats)
library(data.table)
library(ggpubr)
library(scales)
library(plotly)
library(dplyr)
library(vegan)
library(ggplot2)
library(RColorBrewer)
library(ComplexHeatmap)
library(stats)
```

load repertoires
```{r}
lung <- repLoad("~/Desktop/TCR-PBMC")
```


low clones
```{r}
clones<-repExplore(lung$data, .method = "clones")
blood_order = arrange(clones, Clones)
rownames(blood_order) = c()
head(blood_order, 8)
```

omit small samples
```{r}
lung$data[which(names(lung$data) %in% c("MDA-4378-C"))] <- NULL
```

barplot of sample frequency of each stage
```{r}
# create dataframe of each disease stage frequency
#stage_frequency <- data.frame(value = toupper(lung$meta[lung$meta$`Tissue Source`=="PBMC"|lung$meta$`Tissue Source`=="N\\A",]$`Disease Stage`)) %>%
#  count(value)

stage_frequency <- data.frame(value = toupper(lung$meta[lung$meta$`Tissue Source`=="PBMC",]$`Disease Stage`)) %>%
  count(value)
ggplot(stage_frequency, aes(x = value, y = n, fill = value)) +
  geom_bar(stat = "identity") +
  labs(title="Sample frequency by disease stage", x = "Disease stage", y = "Frequency") + 
  scale_color_discrete(name="Stage", labels=c("I","II","III")) +
  scale_fill_grey() 
```


subsampling for more accurate analysis
```{r}
sub <- repSample(.data = lung$data, .method = "downsample")
```

save as RDS
```{r}
saveRDS(sub, "sub-pbmc.Rdata")
saveRDS(lung$meta, "meta.Rdata")
```

load RDS
```{r}
lung <- {}
sub <- readRDS(file="sub-pbmc.Rdata")
lung$meta <- readRDS(file="meta.Rdata")
```

shannon diversity + richness index
```{r}
# shannon index for subsampled samples
shannon_div <- lapply(sub, function(x) {diversity(x$Clones,index = "shannon")})
# shannon index for pre-subsmpled samples
shannon_div <- as.matrix(shannon_div) %>% cbind(sapply(names(sub), function(x){lung$meta[lung$meta$Sample == x,]$`Disease Stage`}))

df_div <- as.data.frame(shannon_div)
colnames(df_div) <- c("shan","stage")
# convert datatypes for plotting later on
df_div$stage <- as.factor(as.numeric(factor(unlist(df_div$stage))))
df_div$shan <- as.double(df_div$shan)
# true false sickness measure for plot comparing
#df_div$is.sick <- grepl("MDA",rownames(df_div))
# richness measure for subsmapled and pre-subsmapled samples
df_div$rich <- as.double(lapply(sub, function(x) {specnumber(x$Clones)}))
df_div$presubrich <- as.double(lapply(lung$data, function(x) {specnumber(x$Clones)}))
```

richness barplot pre subsample
```{r}
ggplot(df_div, aes(x = rownames(df_div), y = presubrich)) +
  geom_bar(stat='identity') + 
  labs(title = "Richness index", x="Samples", y="Richness level") +
  theme_bw() + theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```

richness barplot
```{r}
ggplot(df_div, aes(x = rownames(df_div), y = rich)) +
  geom_bar(stat='identity') + 
  labs(title = "Richness index", x="Samples", y="Richness level") +
  theme_bw() + theme(axis.text.x=element_blank(),
        axis.ticks.x=element_blank())
```


richness index by disease stage
```{r}
ggplot(df_div, aes(x = stage, y = rich, colour = stage)) +
  geom_boxplot() + 
  scale_color_discrete(name="Stage", labels=c("I","II","III","H")) +
  labs(title = "Richness index by disease stage", x="Disease stage", y="Richness level") +
  theme_bw() +
  stat_compare_means()
# anova
TukeyHSD(aov(rich~stage,data=df_div))
```

shannon diversity by disease stage
```{r}
ggplot(df_div, aes(x = stage, y = shan, colour = stage)) +
  geom_boxplot() + 
  scale_color_discrete(name="Stage", labels=c("I","II","III","H")) +
  labs(title = "Shannon diversity index by disease stage", x="Disease stage", y="Diversity level") +
  theme_bw() +
  stat_compare_means() +
  stat_compare_means()
# anova test
TukeyHSD(aov(shan~stage,data=df_div)) 
```


top clonal proportions by disease stage
```{r}
clontop <- repClonality(sub, "top")
vis(clontop, .by = "Disease Stage", .meta = lung$meta)
df_div$clonal10 <- clontop[,"10"]
df_div$clonal3000 <- clontop[,"3000"]

# check if theres any significant group
TukeyHSD(aov(clonal10~stage,data=df_div))
TukeyHSD(aov(clonal3000~stage,data=df_div))
```

repetoire overlap analysis
```{r}
public <- repOverlap(sub, .method = "public", .verbose = F, .col = "aa")
morisita <- repOverlap(sub, .method = "morisita", .verbose = F, .col = "aa")
jaccard <- repOverlap(sub, .method = "jaccard", .verbose = F, .col = "aa")
```

save rds for later
```{r}
saveRDS(morisita, "morisita-pbmc.Rdata")
saveRDS(public, "public-pbmc.Rdata")
saveRDS(jaccard, "jaccard-pbmc.Rdata")
```

load rds
```{r}
morisita <- readRDS(file="morisita-pbmc.Rdata")
public <- readRDS(file="public-pbmc.Rdata")
jaccard <- readRDS(file="jaccard-pbmc.Rdata")
```

display overlap analysis ordered by disease stage
```{r}
# Find the corresponding row indices in the repOverlap matrix for each sample in the metadata
indices <- match(rownames(df_div), rownames(morisita))

# Create a new metadata matrix with matching row indices
metadata_matrix <- df_div[indices, ]

# Order the metadata matrix based on disease stage
ordered_metadata <- metadata_matrix[order(metadata_matrix$stage), ]

# Reorder the repOverlap matrix
reordered_repOverlap_mor <- morisita[order(metadata_matrix$stage), order(metadata_matrix$stage)]
reordered_repOverlap_jac <- jaccard[order(metadata_matrix$stage), order(metadata_matrix$stage)]
reordered_repOverlap_pub <- public[order(metadata_matrix$stage), order(metadata_matrix$stage)]

stage_anno_list <- setNames(as.character(ordered_metadata$stage), rownames(ordered_metadata))


heat_mor <- Heatmap(reordered_repOverlap_mor, column_title = "Morisita overlap ordered by disease stage", cluster_rows = FALSE, cluster_columns = FALSE,
        top_annotation = HeatmapAnnotation(stage_top = stage_anno_list, col = list(stage_top = c("1" = "orange", "2" = "red", "3" = "brown")),
                                        annotation_legend_param = list(stage_top = list(labels = c("I", "II", "III")))),
        left_annotation = rowAnnotation(stage_left = stage_anno_list, col = list(stage_left = c("1" = "orange", "2" = "red", "3" = "brown")),
                                        annotation_legend_param = list(stage_left = list(labels = c("I", "II", "III"))))
        
        )
heat_jac <- Heatmap(reordered_repOverlap_jac, column_title = "Jaccard overlap ordered by disease stage", cluster_rows = FALSE, cluster_columns = FALSE,
        top_annotation = HeatmapAnnotation(stage_top = stage_anno_list, col = list(stage_top = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_top = list(labels = c("I", "II", "III")))),
        left_annotation = rowAnnotation(stage_left = stage_anno_list, col = list(stage_left = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_left = list(labels = c("I", "II", "III"))))
        
        )
heat_pub <- Heatmap(reordered_repOverlap_pub, column_title = "Public overlap ordered by disease stage", cluster_rows = FALSE, cluster_columns = FALSE,
        top_annotation = HeatmapAnnotation(stage_top = stage_anno_list, col = list(stage_top = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_top = list(labels = c("I", "II", "III")))),
        left_annotation = rowAnnotation(stage_left = stage_anno_list, col = list(stage_left = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_left = list(labels = c("I", "II", "III"))))
        
        )
draw(heat_mor)
draw(heat_jac)
draw(heat_pub)
```

display overlap analysis by clustering
```{r}
pub_clust <- Heatmap(public, column_title = "Public overlap ordered by clustering", top_annotation = HeatmapAnnotation(stage_top = stage_anno_list, col = list(stage_top = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_top = list(labels = c("I", "II", "III")))),
        left_annotation = rowAnnotation(stage_left = stage_anno_list, col = list(stage_left = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_left = list(labels = c("I", "II", "III"))))
        
        )
mor_clust <- Heatmap(morisita, column_title = "Morisita overlap ordered by clustering", top_annotation = HeatmapAnnotation(stage_top = stage_anno_list, col = list(stage_top = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_top = list(labels = c("I", "II", "III")))),
        left_annotation = rowAnnotation(stage_left = stage_anno_list, col = list(stage_left = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_left = list(labels = c("I", "II", "III"))))
        
        )
jac_clust <- Heatmap(jaccard, column_title = "Jaccard overlap ordered by clustering", top_annotation = HeatmapAnnotation(stage_top = stage_anno_list, col = list(stage_top = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_top = list(labels = c("I", "II", "III")))),
        left_annotation = rowAnnotation(stage_left = stage_anno_list, col = list(stage_left = c("1" = "pink", "2" = "orange", "3" = "brown")),
                                        annotation_legend_param = list(stage_left = list(labels = c("I", "II", "III"))))
        
        )
draw(pub_clust)
draw(mor_clust)
draw(jac_clust)
```

```{r}
mor_mds <- repOverlapAnalysis(morisita, "mds+kmeans", .k = 5) # 5
jac_mds <- repOverlapAnalysis(jaccard, "mds+kmeans", .k=7) # 7
pub_mds <- repOverlapAnalysis(public, "mds+kmeans", .k=6) # 6
vis(mor_mds, .by = "Disease Stage", .meta = lung$meta)
vis(jac_mds, .by = "Disease Stage", .meta = lung$meta)
vis(pub_mds, .by = "Disease Stage", .meta = lung$meta)
```

generate dataframes: features x observations. 
proportions of public clonotype sharing
```{r}
#lc_seqs <- pubRep(sub, .quant = "prop", .verbose = F, .col = "aa")
#tumor_seqs <- pubRep(sub_t, .quant = "prop", .verbose = F, .col = "aa")
df_seq_lc <- as.data.frame(pubRep(sub, .quant = "prop", .verbose = F, .col = "aa"))
```

remove na and sample column
```{r}
df_seq_lc[is.na(df_seq_lc)] = 0
#lc_in_t[is.na(lc_in_t)] = 0
df_seq_lc$Samples <- NULL
#df_seq_t_wos <- subset(lc_in_t, select = -c(Samples) )
```

transpose dataframe: observations x features
```{r}
tr_lc <- setNames(data.frame(t(df_seq_lc[,-1])), df_seq_lc[,1])
#tr_t <- setNames(data.frame(t(lc_in_t[,-1])), lc_in_t[,1])
```

check how many coulmns are in the df
```{r}
ncol(tr_lc)
```

perform pca on #healthy# (obsolete) + sick
```{r}
pca_all_result <- prcomp(tr_lc, center = T)
```

plot pca
```{r}
ggplot(pca_all_result$x, aes(PC1,PC2,color=df_div$stage)) + 
  geom_point() + 
  scale_color_discrete(name="Stage", labels=c("I","II","III","H"))
```
